Fail Faster

### 2025-03-20 paper/paper.tex
pass done

To support real monadic programming, we'll need to adjust the type of \texttt{'a Gen.t} slightly.
An initial attempt is to try \texttt{type 'a t = int code -> SR.t code -> 'a}. If we strip the \texttt{code} off the result type,
the functions \texttt{return (x : 'a) = fun \_ \_ -> x} and \texttt{bind (g : 'a t) (k : 'a -> 'b t) = fun size random -> k (g size random) size random}
have the proper types for a monad instance. Then, any combinators of type \texttt{'a Gen.t} before simply become \texttt{'a code Gen.t} with this new version.

However, this definition of bind doesn't have call-by value effect semantics, as
discussed in the previous section!  And because the type of \texttt{g size random} is
just \texttt{'a} (not necessarily \texttt{'a code}), we cannot perform the
let-insertion needed to preserve the CBV effects. To solve this problem, we
turn to a classic technique from the multistage programming literature: writing
our staged programs in continuation-passing style \cite{bondorf92}.
\hg{Slow down! This is all very interesting and very technical! Try making each
of these sentences two sentences, add some citations, and maybe spell this out
wiht an example}
\jwc{I could, but i'm just not sure this is critical. In particular, every time I've ever seen this presented in a paper they just drop it and move on. It's too deep in the weeds for it to be worth explaining, too... nobody's going to read this section except staging fans}

@@

We can then redefine our staged generator monad type to be
\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
\jwc{any types that were \texttt{'a Gen.t} before are now \texttt{'a code Gen.t}}.
This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
use the monadic syntactic sugar of our chosen language. Moreover, we also get to maintain the correct
effect ordering: effectful combinators like \texttt{Gen.int} do their \emph{own} let-insertion, ensuring
that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the randomness library.
For example, \lstinline{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
\lstinline{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}.

This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.\footnote{To ensure that the OCaml code is written correctly, we employ a PBT harness to compare its outputs to the corresponding \bq generator over 1,000 random seeds. 
If no discrepancy arises, we conclude that the generators exhibit equivalent behavior.}

@@

Different generator DSLs handle defining recursive generators differently. Some allow recursive generators
to be defined as recursive functions (or in Haskell's case, recursive values), while others
(like \bq) expose a fixpoint combinator to construct recursive generators.
Generator fixpoint combinators are simliar to a usual \texttt{fix : ('a -> 'a) -> 'a} combinator,
except that they operate at monadic type \texttt{fixed\_point : ('a Bq.t -> 'a Bq.t) -> 'a Bq.t}.
Given a step function that takes a ``handle'' to sample from a recursive generator call, it
ties the knot and builds a recursive generator.

In our case, letting programmers define recursive generators as recursive functions is out of the question.
With staged programming, recursion must be handled with care: it is far to easy to accidentally recursively
define an infinite \texttt{code} value and have the program diverge at compile time, when trying to write
a \texttt{code} representing a recursive program.
To this end, we develop a staged recursive generator combinator\footnote{
In reality, we actually have a more general API that allows programmers to define \emph{parameterized}
recursive combinators, of type \texttt{'r code -> 'a code Gen.t}, for any type \texttt{'r}. See Appendix~\jwc{appendix} for details.
}, whose API is shown in Figure~\ref{fig:staged-recursive-generator}.
The code for this combinator can be found in Appendix~\jwc{appendix}.
The recursion API consists of an opaque type \texttt{'a handle}, and a function \texttt{recurse} to perform recursive calls.
Programmers can then define recursive generators by \texttt{fixed\_point}, which ties the recursive knot.
\jwc{don't want to handle GADTs}

>>>>

To support real monadic programming, we'll need to adjust the type of \texttt{'a Gen.t} slightly.
An initial attempt is to try \texttt{type 'a t = int code -> SR.t code -> 'a}. If we strip the \texttt{code} off the result type,
the functions \texttt{return (x : 'a) = fun \_ \_ -> x} and \texttt{bind g k = fun size seed -> k (g size seed) size seed}
have the proper types for a monad instance. Then, any combinators of type \texttt{'a Gen.t} before simply become \texttt{'a code Gen.t} with this new version.

However, this definition of bind doesn't have call-by value effect semantics, as
discussed in the previous section!  And because the type of \texttt{g size seed} is
now just \texttt{'a} (not necessarily \texttt{'a code}), we cannot perform the
let-insertion needed to preserve the CBV effects. To solve this problem, we
turn to a classic technique from the multistage programming literature: writing
our staged programs in continuation-passing style \cite{bondorf92}.
\hg{Slow down! This is all very interesting and very technical! Try making each
of these sentences two sentences, add some citations, and maybe spell this out
wiht an example}
\jwc{I could, but i'm just not sure this is important. It's too deep in the weeds for it to be worth explaining, too... nobody's going to read this section except staging fans}

@@

We can then redefine our staged generator monad type to be
\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
\jwc{any types that were \texttt{'a Gen.t} before are now \texttt{'a code Gen.t}}.
This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
use the monadic syntactic sugar of our chosen language. Moreover, we also get to maintain the correct
effect ordering: effectful combinators like \texttt{Gen.int} do their \emph{own} let-insertion, ensuring
that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the randomness library.
For example, \lstinline{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
\lstinline{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}.
This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.\footnote{To ensure that the OCaml code is written correctly, we employ a PBT harness to compare its outputs to the corresponding \bq generator over 1,000 random seeds. 
If no discrepancy arises, we conclude that the generators exhibit equivalent behavior.}

@@

Different generator DSLs handle defining recursive generators differently. Some allow recursive generators
to be defined as recursive functions, while others
(like \bq and ScalaCheck) expose a fixpoint combinator to construct recursive generators.
Given a step function that takes a ``handle'' to sample from a recursive generator call, it
ties the knot and builds a recursive generator.

In our case, letting programmers define recursive generators as recursive functions is out of the question.
With staged programming, recursion must be handled with care: it is far to easy to accidentally recursively
define an infinite \texttt{code} value and have the program diverge at compile time, when trying to write
a \texttt{code} representing a recursive program.
To this end, we develop a staged recursive generator combinator\footnote{
In reality, we actually have a more general API that allows programmers to define \emph{parameterized}
recursive combinators, of type \texttt{'r code -> 'a code Gen.t}, for any type \texttt{'r}. See Appendix~\jwc{appendix} for details.
}, whose API is shown in Figure~\ref{fig:staged-recursive-generator}.
The code for this combinator can be found in Appendix~\jwc{appendix}.
The recursion API consists of an opaque type \texttt{'a handle}, and a function \texttt{recurse} to perform recursive calls.
Programmers can then define recursive generators by \texttt{fixed\_point}, which ties the recursive knot.
